The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity

### 2023-06-07 paper/markdown/rebuttal_letter.Rmd

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
subtitle: "Rebuttal Letter"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

\newpage

# Editor

<span custom-style = 'reviewer'>Thank you for the timely and well written piece. 
I have received three reviews from exceptional scholars in the field. 
As you can see, there are many strengths to the paper that they note and the publication is timely. 
However, there are also several concerns raised and I am not entirely sure they can all be sufficiently addressed to warrant publication. 
However, I think it is worth a shot. 
Please be attentive to reviewer concerns for a resubmission.</span>

Thank you for appreciation of our manuscript and giving us a chance to address Reviewers' issues. 
First we would like to thank the Reviewers for their very careful, thoughtful and helpful assessment and comments. 
We acknowledge that such detailed reviews take a lot of time and demand a lot of commitment and hope that we have done them justice in the revised manuscript.

Overall, we agree that the initial manuscript version was not sufficiently coherent and clear because of the involvement of a large team of researchers in the study. 
Following your advice and the Reviewers' suggestions, we reworked thoroughly the analysis and the text. 
We would like to point out that this is one of very few manuscripts that characterize activity of the serotonin, kynurenine and catecholamine pathways, i.e. the metabolic system that controls systemic availability of indolamine and catecholamine neurotransmitters in COVID-19 [@Santiago-Mujika2022; @Saito2022; @Lionetto2021; @Bizjak2022; @Dewulf2022; @Ceballos2022]. 
Importantly, their role in development of symptoms of mental health disorders in COVID-19 patients was proposed [@Kucukkarapinar2022; @Bower2022]. 
By analysis of two cohorts, we were able to identify key factors affecting activity of those metabolic pathways and investigate their temporal activity profile in acute and sub-acute SARS-CoV-infection and recovery, which adds to novelty of our work.

We have re-analysed the data to address the reviewers´ issues and have rewritten many parts of the manuscript following the reviewers' suggestions. 
In particular, we now clearly and explicitly state the gap in the literature in __Introduction__ and the analysis endpoints in __Methods__ and __Results__. 
To facilitate interpretation of our results, we now follow a top - bottom analysis approach. 
In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting the key metabolites and activity markers of the kynurenine, serotonin and catecholamine pathways. 
Subsequently, we explore these factors in more detail by correlation and network analysis, and canonical statistical tests. 
Following the Reviewer's suggestion we also keep analysis results in the SIMMUN and INCOV cohort separately for the sake of clarity. 
Concerning the manuscript text we also improved the preciseness of the language and adapted the __Discussion__ part. 
Please see below the point-to-point response for details. 
Please notice that due to the scale of changes in the text, it was not feasible to mark all changes introduced during revision. 
The most relevant changes are <span custom-style = 'revision'>highlighted in red</span>. 
We hope that our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.

@@

# Reviewer 1

## Issue 1

<span custom-style = 'reviewer'>The study assesses the relationship between psychological variables, inflammatory markers, and concentrations of aminoacids in individuals after SARS-CoV-2 infection compared to uninfected controls in two distinct cohorts (SIMMUN; INCOV). 
Moreover, the difference between the control group and different states after SARS-CoV-2 infection (acute, sub-acute, recovery) are compared to the control group within the longitudinal INCOV study cohort.
In addition, the study aims to test the hypothesis that Phenylalanine, Tyrosine, Tryptophan, Kynurenine, and their respective ratios, are predicted by psychological variables inflammation, and SARS-CoV-2 infection-related variables (infection yes vs no, Anti-Antibody titers).
First, it should be mentioned that the graphs are of very high quality and aesthetics and summarize the results very well. 
In addition, the use of a very sophisticated analysis (multiparametric linear modeling) and the detailed presentation of this analysis, and the R packages used are highly creditable. 
However, several weaknesses, including, biochemically wrong assumptions regarding the assessed parameters, lack of transparency regarding the pre-processing of the data, and measurement comparability between the assessed cohorts limit its significance and impact. 
The quality of the paper also suffers from a lack of references, scientifically inadequate language, and multiple spelling and grammatical errors. Because of the many issues that arise, I would not recommend the manuscript for publication in _Psychoneuroendocrinology_.</span>

We thank the Reviewer for the careful assessment of our manuscript and valuable feedback which were very helpful in revising our work. 
First we would like to apologize for imprecise wording, not including enough references and any language errors. 
We have reworked large parts of the manuscript in order to convey our research aims, results and their interpretation in a more clear way. 
For these reasons, we re-formatted the analysis pipeline and the manuscript in the following way:

* we unified nomenclature of the metabolic pathways investigated in our manuscript and term them now 'serotonin pathway', 'kynurenine pathway' and 'catecholamine pathway' in line with prominent review articles [@Badawy2017; @Lukic2022; @Meiser2013]

* we reorganized and rewrote many parts of __Introduction__.

* we also more explicitly and clearly state the ‘gap in the literature’ and our research aims in the last paragraph of __Introduction__.

* we included the __Analysis endpoints__ section of __Methods__. Here we describe our analytic approach, define outcome variables and candidate explanatory factors.

* we show the analysis strategy for the SIMMUN and INCOV cohort in __Supplementary Figure S1__. We hope that this makes it easier to follow our analysis. Additionally we restrict the analyzed SIMMUN and INCOV cohorts to participants with the complete set of outcome and explanatory variables (__Supplementary Figure S1__, __Supplementary Table S1__ and __S3__). This makes the results of particular analyses more easily comparable to one another. Additionally, to investigate if and how exclusion of participant affected the SIMMUN cohort characteristic and if it could be regarded as a source of bias, we compared the excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the cohorts__.

* We excluded CRP and IL6 from analyses presented in the revised manuscript. Both  markers were found to be slightly above their reference ranges (0.5 mg/dL and 7 pg/mL) solely in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of SIMMUN individuals, respectively. They are hence unlikely to be biologically and clinically relevant predictors of the kynurenine and catecholamine pathway activity. Details on this are now described in __Supplementary Methods__.

* We provided information on all analyzed variables of the SIMMUN (__Supplementary Table S1__) and INCOV cohort (__Supplementary Table S3__) with their format, units and, if applicable, normality-stabilizing transformations.

* In order to improve clarity of __Results__, we have now reordered and reformatted it. We now present the results in a ‘top - bottom’ approach. In brief, both in the SIMMUN and the INCOV cohort, we employ multi-parameter modeling (linear regression or robust linear regression), to identify key explanatory factors significantly affecting the outcome. Subsequently, we investigate those significant predictors in more detail by ‘univariable’ correlation analysis and statistical hypothesis testing. This analysis strategy has recently been used in two COVID-19-related publications of our group [@Hufner2022; @Hufner2023].

@@

## Issue 2

<span custom-style = 'reviewer'>__Introduction:__ In the introduction the term "psychoneuroimmunological mechanisms" is mentioned twice. 
I would appreciate it if at least some of the mechanisms which are relevant to the paper are specified.</span>

As outlined in __Issue 1__, we re-ordered and re-wrote the entire __Introduction__ section. Following discussions in the core author team, we decided to replace the general term ‘psychoneuroimmunological’ with more direct descriptions of the mechanisms we are referring to i.e. the role of inflammation on IDO activity are the effect of glucocorticoids on TDO activity. We have also changed the title of the manuscript.

@@

## Issue 3

<span custom-style = 'reviewer'>__Introduction:__ Some statements made in the introduction are without reference, very general, and the conclusions weakly justified. 
For instance, TRP, KYN, and KYN/TRP are listed as markers of serotonin availability. 
However, the available evidence for the relationship between increased degradation of TRP to KYN and its effects on serotonin availability is controversial as only 1-2% of tryptophan metabolism goes into the serotonin pathway (https://doi.org/10.3390/metabo12060514). 
In addition, line 60 mentions a depressiogenic and an anxiogenic effect of QUIN, and an anxiety- and depression-amplifying effect of kynurenine and catabolites is stated in line 9 (page 3). 
Should this statement refer to existing cross-sectional findings (e.g., higher QUIN in patients with MDD compared to healthy controls) it is very generalized and not sufficiently justified. 
If the statement refers to evidence that suggests a causal relationship, a reference is urgently needed.</span>

We agree that the role of the kynurenine pathway metabolites as well as systemic, i.e. peripheral availability of tryptophan, serotonin and dopamine in mental health are controversial, not sufficiently supported by direct experimental evidence and definitely needs a critical view on the topic [@Marx2020; @Brown2021; @Moncrieff2022; @Goldstein1999]. 
We find that one of the many challenges is that many mechanistic studies are done in the animal model, making the transfer to the human situation questionable [@Brown2021]. 
On the other hand human studies are quite equivocal and, while there are influential theoretical manuscripts on the issue, surprisingly few original research has been published so far [@Brown2021]. 
To address you issue, we decided to give a more balanced view on the serotonin, kynurenine, catecholamine pathways and their role in mental health while still keeping the text relatively compact. 
We now give a more clear, but short introduction to there three biochemical pathways in the introduction section. 
We provide an ample discussion of the possible role of these pathways in linking mental health and physical conditions. 
In __Discussion__ and __Limitations__ we comment on controversy over the role of systemic serotonin, kynurenine or catecholamine pathway in depression and other psychiatric disorders [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999].

Additionally, to address the question how TRP levels may affect systemic serotonin pathway activity in COVID-19 patients, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursors and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age. 
As presented in __Figure 2__, we could indeed find a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection time course. 
In a more detailed correlation and network analysis of the INCOV dataset (__Figure 4A__, __Supplementary Figure S6__), we could observe significant, weak-to-moderate strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection. 
This suggests that systemic serotonin synthesis may be limited by blood TRP availability to some extent during SARS-CoV-2-dependent inflammation, which was characterized by lowered plasma TRP concentration and elevated KYN levels.

@@

## Issue 4

<span custom-style = 'reviewer'>__Introduction:__ The introduction sentence of the last paragraph in the introduction seems not adequate for the following hypotheses. 
It highlights that low-grade inflammation (as I understand it, in the case of this study IL-6 and CRP concentrations) is associated with mental health. 
However, the presented hypotheses use IL-6 and CRP and mental stress, anxiety, and depressive symptoms as predictors in their analysis (multi-parameter linear modeling) and TRP, KYN, KYN/TRP, PHE, TRY, PHE/TRY as dependent variables.</span>

We are sorry for this unclarity. 
In the revised manuscript, we formulate the aims, and define outcomes and explanatory factors. 
Please refer to the last paragraph of __Introduction__ and the subsection __Analysis endpoints__ of __Methods__.

@@

## Issue 5

<span custom-style = 'reviewer'>__Methods:__ while the SIMMUN cohort is described in detail, basic information about the INCOV cohort is missing and should be provided.</span>

We agree with the Reviewer that the INCOV cohort deserves a more thorough description. 
We included short methodological information of the INCOV cohort in __Methods/Procedures__ and __Supplementary Methods__ [@Su2020; @Su2022].

@@

As specified now in __Methods/Procedures__, values beyond or at their detection limits were substituted with the lower or upper detection limit. 
CRP and IL6 were removed from the revised analysis pipeline for the reasons specified in __Issue 1__.

## Issue 8

<span custom-style = 'reviewer'>__Methods:__ Kruskal-Wallis test with Man-Whitney U tests are applied to assess the differences be-tween uninfected individuals and infected individuals at different time points. 
While the pairwise comparisons are adequate for this research question, the overall statistic of the Kruskal-Wallis test is not useful in this context, as it tests whether there is an over-all difference between independent groups. 
I would therefore suggest using a linear model with predefined contrasts (uninfected vs. acute; uninfected vs. sub-acute; uninfected vs. recovery) inclusive control for multiple tests. 
If the assumptions of linear regression are not fulfilled, robust methods can be used (e.g., R-package: MASS; https://www.rdocumentation.org/packages/MASS/versions/7.3-58.2/topics/rlm). 
Additionally, the correlations between the assessed dependent variables seem rather high, why controlling for multiple comparisons using Benjamini & Yekutieli procedure (2001) or even robust MANOVAS (see: https://cran.r-project.org/web/packages/MANOVA.RM/vignettes/Introduction_to_MANOVA.RM.html) could be also necessary. Additionally, the terms "acute", "sub-acute and "recovery" should be described in the notes of __Figure 3__.</span>

We are grateful for these suggestions and included them in our analyses. 
In the revised manuscript, we applied robust linear regression with the MM algorithm and Huber's psi function [@Huber2011; @Ripley2022] for multi-parameter modeling (__Figure 2__) and time-course modeling of the INCOV dataset (__Figure 3__). 
Additionally, we specify in legends of figures with INCOV cohort data the median time interval between diagnosis and sampling for the acute, sub-acute and recovery timepoints. 
As suggested, we applied multiple testing adjustment throughout the analysis pipeline. 
Except for multi-parameter modeling, p values were adjusted separately for each analysis step (e.g. correlation analysis, time-course modeling) with the false discovery rate method known also as 'Benjamini-Hochberg' [@Benjamini1995].

@@

## Issue 9

<span custom-style = 'reviewer'>__Psychometric assessment:__ In the description of the psychometric assessment, the source of the HADS is missing as well as a calculation of the reliability of the used questionnaires for the SIMMUN study cohort.</span>

We apologize for the missing references and cite the seminal HADS paper and a report dealing with the clinically relevant HADS cutoffs [@Bjelland2002; @Zigmond1983] in the revised text. 
We also checked consistency of the PSS-4 and HADS psychometric tools employed in our report. 
Since the HADS scales, and in particular, the HADS depression tool were found to be non-tau-equivalent by factors analysis (__Supplementary Figure S2A__), we decided use McDonald's $\omega$ as a consistency measure instead of the more popular Cronbach's $\alpha$ [@Revelle2022; @McDonald1999]. 
The PSS-4 tool demonstrated a good and the HADS scales an excellent consistency, as presented in __Supplementary Figure S2B__.

@@

## Issue 10

<span custom-style = 'reviewer'>__Results:__ First, it would be useful to distinguish between cohorts in the descriptive characterization of cohorts. 
Furthermore, concrete numbers should be given instead of rough descriptions (e.g., "about half" (page 8, line 19)). 
Second, assessing the effect as "dramatic" is not appropriate in a scientific context. 
If at all, existing assessments based on effect sizes can be made using generally accepted definitions (e.g., according to Cohen).</span>

We concur with the Reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S5__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
The cohorts differed in age and sex distribution, rates of SARS-CoV-2 infection and infection severity. 
In particular, most INCOV study participants experienced moderate-to-critical COVID-19 and required hospitalization. 
This may explain the stronger effects of inflammation and infection timepoint on inflammatory cytokines and the metabolites of interest, which we discuss in __Discussion__. 
In the revised text, Figures and Tables we also present effect size metrics for statistical hypothesis tests, correlation and modeling. 
We interpret them as proposed in the handbook by Cohen [@Cohen2013] (__Supplementary Methods__).

@@

## Issue 11

<span custom-style = 'reviewer'>__Results:__ In general, numbers (test statistics, p-values) are missing in the results part and should be added. Additionally, effect sizes should be presented.</span>

We present now p values and effect sizes or model $\beta$ with 95% confidence intervals, as appropriate, for the most important findings in Abstract and Results. 
Yet, for the sake of readability, we were not able to provide significance and effect size information for all effects. 
In particular, we do not specify p values and effect sizes for non-significant effects and state instead clearly that the effect missed statistical significance. 
We abstained from referencing test statistics such as $\chi^2$ or $t$, since they depend on the number of complete observations and hence do not deliver reliable information on the effect size.

@@

Following your suggestion, we avoid this wording in the revised manuscript. 
Please refer to __Issue 11__ for details.

## Issue 13

<span custom-style = 'reviewer'>__Results:__ The terminology of strata (page 10, line 10) can be misleading as no stratification was performed. Probably it would be better to use the term group or subgroup.</span>

Following your suggestion, we have replaced the term ‘strata’ in the revised text by 'subgroup' or 'subset'.

@@

## Issue 14

<span custom-style = 'reviewer'>__Introduction:__ The use of subheadings would be helpful to facilitate the understanding of the first part of the introduction.</span>

Since we re-wrote and shortened the __Introduction__ section, we abstained from using subheadings; such format is also not common in original research manuscripts in _Psychoneuroendocrinology_. 
To facilitate interpretation, each topic of __Introduction__ is described in a separate paragraph. 
In turn, we employ subheadings in __Discussion__.

@@

## Issue 15

<span custom-style = 'reviewer'>__Results:__ The statement that TRP is a precursor of IDO is not correct!</span>

Thanks for the attentive assessment, this misleading wording was removed. 
Please refer to the metabolic pathway nomenclature in __Issue 1__.

@@

For the sake of clarity, we present characteristic of the SIMMUN and INCOV cohorts in separate tables (__Table 1__ and __Table 2__).

## Issue 18

### 2023-06-07 paper/markdown/main_manuscript.Rmd
Analysis and manuscript files with feedback
Author's feedback included in the analyses and manuscript files. Effect of gender and BMI included in the INCOV cohort multi-parameter models. Changes and shortening of the manuscript text.

---
title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

csl: frontiers-in-immunology.csl

@@

@@

@@

@@

The SIMMUN study was conducted in accordance with the Declaration of Helsinki and European Data Policy. 
All participants gave written informed consent prior to enrollment. 
Participants' data were stored and analyzed in anonymized form.
This study was approved by the ethics committee of the Medical University Innsbruck, Austria (approval number: 1132/2020). 
No approval by the ethics committee was required for analysis of the published anonymized INCOV data set [@Su2022].

@@

Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck Medical University (Innsbruck, Austria) and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study. 
The study enrollment was conducted between 10. June 2020 (first patient enrolled) and 27.May 2021 (last study visit). 
The inclusion criteria were age of 18 - 70 years, proficiency in German language, residence in the study region (Tyrol, Austria), results of a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck. 
The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, or acute or chronic inflammatory illness and treatment with oral corticosteroids. 
The analysis inclusion criterion was the complete study variable dataset (__Supplementary Table S1__). 
A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were included in the analysis (__Figure \@ref(fig:fig-inclusion)__, __Table \@ref(tab:tab-stigma-cohort)__). 
Significant differences between the analyzed participants and participants excluded due to data missingness are listed in __Supplementary Table S2__.

@@

Proteome and metabolome data and clinical information for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 
In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study participants with the complete age information and dataset of metabolites related to serotonin and dopamine turnover and major inflammatory cytokines were included (__Figure 1__, __Table \@ref(tab:tab-incov-cohort)__, __Supplementary Table S3__).

@@

Details on study procedures are provided in __Supplementary Methods__.

@@

The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
The SIMMUN study data were gathered during in-person visits including a physician assessment, supervised completion of self-rating questionnaire and a blood sample collection.

Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed mental illness, self-reported chronic somatic conditions, smoking and alcohol consumption history, result and date of the SARS-CoV-2 PCR test were surveyed during the study visit or extracted from electronic patient records.

Inflammatory markers: plasma neopterin concentrations (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck. 
Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, and the KYN/TRP and PHE/TYR ratios calculated [@Capuron2011; @Widner1997; @Neurauter2008]. 
Plasma titer of immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) were quantified by ELISA as described before [@Deisenhammer2021]. 
Laboratory measurements at and beyond the detection limits were substituted with the lower or upper detection limit value, respectively.

Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
The total possible score range for each subscale is 0 to 21, with higher scores indicating more intense symptoms of anxiety/depression. 
Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points [@Bjelland2002; @Zigmond1983].

In order to improve normality of some numeric study variables prior to linear modeling and statistical hypothesis testing with parametric tools, logarithm or square root transformations were applied (__Supplementary Table S1__)

@@

Details of data transformation and analysis are provided in __Supplementary Methods__.

@@

>>>>

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author:
  - Katharina Hüfner:
      institute: [psych2]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true
  - Sophia Vedova:
      institite: [psych2]
  - Piotr Tymoszuk:
      institute: [daas]
  - Philipp Nelles:
      institute: [psych2]
  - Tobias Bruckner: 
      institute: [psych2]    
  - Eberhard Deisenhammer: 
      institute: [psych2, psych1]
  - Jonas Egeter: 
      institute: [psych2]
  - Alexander Egger: 
      institute: [zimcl]
  - Matyas Galffy: 
      institute: [psych2]
  - Johannes Giesinger: 
      institute: [psych2]
  - Jens Lehmann: 
      institute: [psych1]
  - Maria Oberhammer: 
      institute: [psych1]
  - Joachim Rockenschaub: 
      institute: [psych2]
  - Magdalena Sacher: 
      institute: [surg]
  - Bernhard Holzner: 
      institute: [psych2]
  - Johanna M Gostner: 
      institute: [bio]
  - Barbara Sperner-Unterweger: 
      institute: [psych2]

institute:
  - psych2: University Hospital of Psychiatry II, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria. 
  - psych1: University Hospital of Psychiatry I, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
  - surg: Department of Visceral, Transplant and Thoracic Surgery (VTT), Medical University Innsbruck, Innsbruck, Austria. 
  - zimcl: Central Institute of Clinical and Chemical Laboratory Diagnostics, University Hospital of Innsbruck, Innsbruck, Austria.
  - bio:  Insitute of Medical Biochemistry, Medical University Innsbruck, Innsbruck, Austria

@@

csl: psychoneuroendocrinology.csl

@@

Running title: Activity of the kynurenine, serotonin and dopamine metabolic pathways and COVID-19

# Abstract

__Background:__ COVID-19 patients frequently develop symptoms of depression, anxiety and mental stress. 
The serotonin, kynurenine and catecholamine metabolic pathways determine systemic levels of serotonin and dopamine, whose insufficiency was linked to mental health disorders. 
We investigated factors that affect their activity in SARS-CoV-2 infection and recovery.

__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were analyzed. 
Demographic and clinical characteristic, inflammatory markers, SARS-CoV-2 infection, signs of depression and anxiety (HADS), and mental stress (PSS-4) served as explanatory variables. 
Plasma markers of the serotonin and kynurenine (serotonin, kynurenine/tryptophan ratio), and catecholamine pathway activity (dopamine 3-O-sulfate, phenylalanine/tyrosine ratio) were modeled by multi-parameter linear regression.

__Results:__ In the SIMMUN cohort, the inflammatory marker 
neopterin ($\beta$ = `r my_beta(mod_eli$inference$kyn_trp, 'log_neo')`), 
SARS-CoV-2-positivity (`r my_beta(mod_eli$inference$kyn_trp, 'infection', show_ci_txt = FALSE)`), 
mental stress (`r my_beta(mod_eli$inference$kyn_trp, 'pss_stress_score', show_ci_txt = FALSE)`) and age were positively associated with elevated kynurenine/tryptophan ratio. 
Phenylalanine/tyrosine ratio was lower in 
SARS-CoV-2-positive participants (`r my_beta(mod_eli$inference$phe_tyr, 'infection', show_ci_txt = FALSE)`). 
In the INCOV cohort, inflammation markers were associated with decreased serotonin 
(IL6: `r my_beta(incov_neuro$inference$serotonin, 'IL6_INF', show_ci_txt = FALSE)`) and dopamine 3-O-sulfate 
(interferon-gamma: `r my_beta(incov_neuro$inference$dopamine, 'IFNG_INF', show_ci_txt = FALSE)`). 
Serotonin (`r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery', show_ci_txt = FALSE)`) and 
dopamine 3-O-sulfate levels (`r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery', show_ci_txt = FALSE)`) were higher during recovery than in acute SARS-CoV-2 infection.

__Conclusion:__ SARS-CoV-2 infection, inflammation, mental stress and age are the key independent predictors of high kynurenine pathway activity, which decreases systemic serotonin levels by depleting its precursor tryptophan. 
These factors are also associated with suppressed catecholamine pathway activity. 
This lowered serotonin and dopamine availability may contribute to mental health problems in COVID-19 patients.

# Keywords

serotonin and dopamine availability, COVID-19, kynurenine, tryptophan, inflammation, mental health

@@

The immune system and the brain interact at multiple levels [@Dantzer2008]. 
Physical conditions such as infections, cardiovascular disease, frailty, autoimmune illness or malignancy are often accompanied by symptoms of depression, anxiety, mental stress and other psychiatric disorders [@Mitchell2011; @Dickens2001; @Mayerl2020; @Rajan2020; @Renault1987; @Hufner2015; @Hufner2022]. 
Inflammation interferes with key metabolic processes controlling serotonin and dopamine availability, the serotonin, kynurenine and catecholamine pathways, and was proposed as an important link between mental and physical health [@Bonaccorso2002; @Gostner2020; @Valdiglesias2018; @Dantzer2008; @Hufner2015; @Brown2021].

Serotonin is synthesized from the essential aminoacid tryptophan (TRP) via serotonin pathway reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
Most systemic TRP is catabolized by the kynurenine pathway, which limits serotonin pathway activity [@Lukic2022]. 
Its first rate-limiting step, the TRP - kynurenine (KYN) breakdown catalyzed by tryptophan 2,3-dioxygenase (TDO) or indoleamine 2,3-dioxygenases (IDO1, IDO2), controls TRP levels available for serotonin synthesis [@Lukic2022; @Badawy2017]. 
Importantly, IDO1 is stimulated by inflammatory signals [@Badawy2017; @Robinson2003]
and TDO activity is increased by glucocorticoids released upon depression and mental stress [@Knox1951; @Maes1990]. 
Kynurenine pathway metabolites such as KYN or quinolinic acid [QUIN] can additionally pass the blood-brain barrier, interfere e.g. glutamatergic signaling and exert neurotoxicity [@Brown2021; @Schwarcz2012]. 
Hence, the kynurenine pathway, whose activity can be assessed by the ratio of blood KYN to TRP (KYN/TRP), poses a connex of inflammatory and hypothalamic–pituitary–adrenal axis signals, systemic serotonin availability and neural signal transmission [@Badawy2017; @Brown2021; @Schwarcz2012]. 
Consequently, elevated levels of circulating kynurenine pathway metabolites, high KYN/TRP, decreased TRP or lowered serotonin were reported for numerous physical conditions and associated with symptoms of mental disorders [@Cervenka2017; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Capuron2011].

Dopamine synthesis via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH). 
PAH, TH and the serotonin pathway enzyme TPH rely on tetrahydrobiopterin (BH4) as an essential cofactor [@Meiser2013; @Neurauter2008]. 
During inflammation, BH4 is oxidized to its inactive derivative and consumed in a competitor reaction of nitric oxide formation [@Rahimian2022a; @Neurauter2008]. 
As a result of such functional BH4 deficiency, reflected e.g. by an elevated blood PHE to TYR ratio, biosynthesis of dopamine and serotonin is suppressed [@Geisler2013; @Vancassel2018; @Capuron2011]. 
Accordingly, changes in PHE and TYR levels have been reported in different physical disorders such as cancer, infections and inflammatory conditions and associated with depression and anxiety [@Vancassel2018; @Geisler2013; @Hufner2015; @Capuron2011].

SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
Apart from sustained physical disability, psychiatric disorders in millions of patients amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022]. 
Increased kynurenine pathway activity suggestive of lowered systemic serotonin availability has recently been identified in acute COVID-19 and during COVID-19 recovery and associated with infection severity [@Santiago-Mujika2022; @Lionetto2021; @Bizjak2022; @Dewulf2022]. 
Theoretical considerations [@Bower2022] and preliminary experimental data suggest that highly active kynurenine pathway along with suppression of serotonin and catecholamine biosynthesis may contribute to symptoms of mental stress, anxiety and depression in COVID-19 patients [@Matits2023; @Kucukkarapinar2022].

Still, clinical and experimental evidence for interaction of inflammation, mental health and systemic availability of serotonin and dopamine in COVID-19 is scarce. 
To address this question, we explored effects of demographic, clinical, psychometric (anxiety, depression, mental stress), inflammation- and SARS-CoV-2-related factors on markers of systemic activity of the serotonin, kynurenine and catecholamine pathways in two cross-sectional cohorts of SARS-CoV-2 patients: the SIMMUN cohort and the published longitudinal INCOV study [@Su2022].

@@

Details on study cohorts, procedures and analysis are provided in __Supplementary Methods__.

@@

The SIMMUN study was conducted in accordance with the Declaration of Helsinki and European Data Policy. 
All participants gave written informed consent prior to enrollment. 
Participants' data were stored and analyzed in anonymized form.
This study was approved by the ethics committee of the Medical University Innsbruck (Austria, approval number: 1132/2020). 
No approval by the ethics committee was required for the published INCOV dataset [@Su2022].

@@

Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR were invited to participate. 
The study was conducted between June 2020 and May 2021. 
The inclusion criteria were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR conducted at the study site. 
The exclusion criteria were active SARS-CoV-2 infection (< 14 days after diagnosis), pregnancy, active malignancy, organ transplantation, surgery in the past 3 months, inflammatory illness and oral corticosteroid treatment. 
Only individuals with complete explanatory and response variables presented in __Supplementary Table S1__ and __Supplementary Figure S1__ were analyzed. 
Significant differences between the analyzed and excluded participants are listed in __Supplementary Table S2__.

@@

Proteome, metabolome and clinical data are [publicly available](https://data.mendeley.com/datasets/96v329bg7g/1) [@Su2022]. 
Samples with complete explanatory and response variables presented in __Supplementary Figure S1__ and __Supplementary Table S3__ were analyzed.

@@

@@

The SIMMUN data were collected during a single on-site visit at median 139 days after the SARS-CoV-2 PCR (interquartile range: 119 - 157) or extracted from electronic patient records (__Supplementary Table S1__). 
Demographic and clinical characteristic included age, sex, body mass index (BMI), mental disorders diagnosed by a physician, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR. 
Inflammatory markers, plasma neopterin concentrations (NEO) measured by ELISA [@Werner1990] and neutrophil/lymphocyte ratio (NLR) were determined by a certified laboratory at the University Hospital of Innsbruck. 
Plasma TRP, KYN, PHE and TYR were determined by high-performance liquid chromatography [@Widner1997; @Neurauter2008]. 
Immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) was quantified by ELISA [@Deisenhammer2021]. 
Laboratory measurements beyond the detection limits were substituted with the respective detection limit value. 
Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Clinically relevant symptoms of anxiety and depression were defined as $\geq$ 8 points in the hospital anxiety and depression scale (HADS) [@Bjelland2002; @Zigmond1983].

@@

@@

## Characteristic of the cohorts

Two cohorts of uninfected and SARS-CoV-2-infected individuals were analyzed (SIMMUN and INCOV). 
Out of `r nrow(stigma$data)` individuals enrolled in the SIMMUN study, 
`r nrow(stigma$analysis_tbl)` participants with complete study variables were analyzed (__Supplementary Figure S1__, __Supplementary Table S1__). 
The excluded individuals were characterized by more frequent mental disorders and clinically relevant symptoms of depression and anxiety (HADS), higher scores of mental stress (PSS-4) and less frequent SARS-CoV-2 infections than the analyzed participants (__Supplementary Table S2__). 
In the analyzed SIMMUN cohort, males represented 
`r my_percent(cohort$analysis_tbl$stigma, 'sex')['male']`% and the median age was 
`r median(cohort$analysis_tbl$stigma$age)` years. 
